Fractional Types

### 2019-11-20 fracGC/p.tex
revx
done with sec 4 for now

To make this idea work, several technical details are needed. Most
notably, we must exclude the empty types from this creation and
annihilation process. Otherwise, we would be able to prove that:
\[\begin{array}{rcll}
1 &=& 0 * 1/0 & \textrm{by~} \eta_0 \\
&=& 0 & \textrm{by~} \absorbr
\end{array}\]
We therefore extend the syntax of core $\Pi$ in Sec.~\ref{sub:core} as follows:
\[\begin{array}{lrcl}
\textit{Value types} & t &::=& \cdots \alt \oneover{t} \\
\textit{Non-empty types} & \net \\
\textit{Values}      & v &::=& \cdots \alt \gcv \\
\textit{Level-1 types} &&& t \leftrightarrow t \\
\textit{Level-1 programs} & c &::=& \cdots \alt
   \eta : 1 \leftrightarrow (\net * \oneover{\net}) \\
   &&& ~~~~~\alt \epsilon : (\net * \oneover{\net}) \leftrightarrow 1 \\
\textit{Level-2 types} &&& c \Leftrightarrow c \\
\textit{Level-2 programs} & \alpha &::=& \cdots 
  \alt \mathit{inv} : \eta \fatsemi \epsilon \Leftrightarrow \idc 
\end{array}\]

The metavariable $\net$ excludes all types that are equivalent to the
empty type, e.g., $(1 + 1) * 0$. For now, all gc values are represented
at runtime using a trivial value denoted by $\gcv$. We also explicitly
add the coherence condition on the creation and annihilation process
as a level-2 program, thus enforcing the duality $\eta$ and
$\epsilon$.

@@

@@

\vspace*{-0.5cm}
$\vdots$

\EtaEpsilonEval{}

When ancilla are created they are created with a specific
value. This value is encoded in the circuit. The value of fractional
type represents the ``garbage collector.''  The garbage collector is
specialized to collect a single value, which is the value used to
create the ancilla value. At this point, if the garbage collector
encounters a different value, it doesn't know what to do, and would
need to throw an exception. This situation in \verb|PiFracDyn| is
similar to current solutions (e.g. Quipper) but actually more general
as it allows non-scoped ancilla allocation and de-allocation. Show
examples.

@@

The first example below illustrates the basic functionality of
ancillae. The Agda code is written in a style that reveals the
intermediate steps for convenience. The circuit has one input and one
output. Immediately after receiving the input, the circuit generates
an ancilla wire and its corresponding gc process (first two steps in
the Agda definition). The original input and the ancilla wire interact
using two \AgdaFunction{CNOT} gates, after which the ancilla wire is
redirected to the output (next three steps in the Agda code). Finally
the original input is gc'ed (last two steps in the Agda code). The
entire circuit is extensionally equivalent to the identity function.

@@

\begin{tikzpicture}[x=0.75pt,y=0.75pt,scale=0.8, yscale=-1]
%uncomment if require: \path (0,300); %set diagram left start at 0, and has height of 300
%Straight Lines [id:da9303973632982276] 
\draw    (40,80) -- (190,80) ;
%Shape: Arc [id:dp2904677718814479] 
\draw  [draw opacity=0] (38.49,110.13) .. controls (28.3,109.22) and (20.27,102.7) .. (20.15,94.89) .. controls (20.02,86.57) and (28.89,79.93) .. (40,80) -- (40.38,95.12) -- cycle ; \draw   (38.49,110.13) .. controls (28.3,109.22) and (20.27,102.7) .. (20.15,94.89) .. controls (20.02,86.57) and (28.89,79.93) .. (40,80) ;
%Straight Lines [id:da32172184710713914] 
\draw    (38.49,110.13) -- (250,110) ;
%Shape: Arc [id:dp8398868072937593] 
\draw  [draw opacity=0] (248.77,109.75) .. controls (262.77,109.82) and (274.32,103.33) .. (274.64,95.17) .. controls (274.95,87.02) and (263.95,80.28) .. (250,80) -- (249.05,94.87) -- cycle ; \draw   (248.77,109.75) .. controls (262.77,109.82) and (274.32,103.33) .. (274.64,95.17) .. controls (274.95,87.02) and (263.95,80.28) .. (250,80) ;
%Straight Lines [id:da7474565765110813] 
\draw    (40.09,30.3) -- (190.09,30.3) ;
%Shape: Arc [id:dp593084478238281] 
\draw  [draw opacity=0] (38.58,60.43) .. controls (28.39,59.52) and (20.36,53) .. (20.23,45.19) .. controls (20.11,36.87) and (28.98,30.23) .. (40.09,30.3) -- (40.47,45.42) -- cycle ; \draw   (38.58,60.43) .. controls (28.39,59.52) and (20.36,53) .. (20.23,45.19) .. controls (20.11,36.87) and (28.98,30.23) .. (40.09,30.3) ;
%Straight Lines [id:da5682447818267274] 
\draw    (38.58,60.43) -- (250.09,60.3) ;
%Shape: Arc [id:dp8936250935910023] 
\draw  [draw opacity=0] (248.86,60.05) .. controls (262.86,60.12) and (274.41,53.63) .. (274.72,45.47) .. controls (275.03,37.32) and (264.04,30.58) .. (250.09,30.3) -- (249.13,45.17) -- cycle ; \draw   (248.86,60.05) .. controls (262.86,60.12) and (274.41,53.63) .. (274.72,45.47) .. controls (275.03,37.32) and (264.04,30.58) .. (250.09,30.3) ;
%Straight Lines [id:da9423433355546923] 
\draw    (190,30) -- (250,80) ;
%Straight Lines [id:da8064437888434692] 
\draw    (190,80) -- (250.09,30.3) ;
\end{tikzpicture}

@@

% of course need de-allocation; keep track of lifetime
% easy: stack but not general
% for general; programmer control or gc

% context of reversible; similar; need allocation; same tradeoffs for
% de-allocation; stack or programmer or gc BUT with caveat. value must
% be restored to its initial value; no information leakage which would
% destroy quantum state

% but reversible also gives you more guarantees; no creation or deletion
% of information means no need to keep track of lifetime of individual
% locations; just need to keep track of ``information''

>>>>

To make this idea work, several technical details are needed. Most
notably, we must exclude the empty types from this creation and
annihilation process. Otherwise, we would be able to prove that:
\[\begin{array}{rcll}
1 &=& 0 * 1/0 & \textrm{by~} \eta_0 \\
&=& 0 & \textrm{by~} \absorbr
\end{array}\]
The second important detail is to ensure that the gc process is
specialized to collect a particular value. We however defer this point to
the next section and for now create gc processes that might be applied
to incorrect values throwing exceptions instead of collecting them. We
therefore extend the syntax of core $\Pi$ in Sec.~\ref{sub:core} as
follows:

\[\begin{array}{lrcl}
\textit{Value types} & t &::=& \cdots \alt \oneover{t} \\
\textit{Non-empty types} & \net \\
\textit{Values}      & v &::=& \cdots \alt \gcv \\
\textit{Level-1 types} &&& t \leftrightarrow t \\
\textit{Level-1 programs} & c &::=& \cdots \alt
   \eta : 1 \leftrightarrow (\net * \oneover{\net}) \\
   &&& ~~~~~\alt \epsilon : (\net * \oneover{\net}) \leftrightarrow 1 \\
\textit{Level-2 types} &&& c \Leftrightarrow c \\
\textit{Level-2 programs} & \alpha &::=& \cdots 
  \alt \mathit{inv} : \eta \fatsemi \epsilon \Leftrightarrow \idc 
\end{array}\]

The metavariable $\net$ excludes all types that are equivalent to the
empty type, e.g., $(1 + 1) * 0$. For now, all gc values are
represented at runtime using a trivial value denoted by $\gcv$. We
also explicitly add the desired coherence condition on the creation
and annihilation process as a level-2 program that would enforce the
duality of $\eta$ and $\epsilon$. At this point, an application of
$\eta$ following by $\epsilon$ might however throw an exception.

@@

\noindent where the operations take an implicit argument asserting the
given type is non-empty. The operational semantics of the language
interprets each combinator (except $\eta$ and $\epsilon$) of type
$t_1 \leftrightarrow t_2$ as a (reversible) function mapping values of
type $t_1$ to values of type $t_2$ (and vice-versa). Extending this
interpreter to handle $\eta$ and $\epsilon$ requires at this point
that the interpreter returns either a value of the appropriate or an
exception:

@@

When ancillae are created they are created with a specific default
value depending on the type. When the gc process encounters a value to
be collected, it checks whether this value is the default value for
the type and if so collects it. Otherwise it returns
\AgdaInductiveConstructor{nothing} indicating an exception.

@@

The first example below illustrates the basic functionality of
ancillae. The Agda code is written in a style that reveals the
intermediate steps for easier correspondence with the figure. The
circuit has one input and one output. Immediately after receiving the
input, the circuit generates an ancilla wire and its corresponding gc
process (first two steps in the Agda definition). The original input
and the ancilla wire interact using two \AgdaFunction{CNOT} gates,
after which the ancilla wire is redirected to the output (next three
steps in the Agda code). Finally the original input is gc'ed (last two
steps in the Agda code). The entire circuit is extensionally
equivalent to the identity function but it does highlight an important
functionality beyond scoped ancillae management: the allocated ancilla
bit is redirected to the output and a completely different bit (with
the proper default value) is collected instead.

@@

The second example illustrates the manipulation of gc processes. A
process for collecting a pair of values can be decomposed into two
processes each collecting one of the values (and vice-versa):

@@
